---
title: "RTMLE data example"
date: "18-04-2024"
output: pdf_document
---

This document uses simulated data to demonstrate the use of the
software for an LTMLE analysis. Note, that this example uses the same
data preparation steps that were used during the Reddie Course. The
example therefore works with several files that contain data and uses
splitting of time dependent variables followed by aggregation of
information in selected time periods to prepare the data.

The hope is that the real data are in similar format or can be brought
into similar format, representing baseline covariates, time-varying
covariates, baseline and time-varying treatment, and outcome,
competing risks and censoring. The example below simulates data where
a treatment is is given that reduces the incidence of events.

```{r, echo=FALSE}
library(heaven) # devtools::install_github("tagteam/heaven")
library(data.table)
# library(ltmle)
library(rtmle)
library(targets)
library(ggplot2)
# source functions
# tar_source("/home/ctp/github/TMLE_for_breakfast/Ltmle/R/")
```
# Step 1 - Get exercise data

## Baseline data and outcomes

The example has a single baseline variable (Age), Time (of event),
Event_Type (0=censor, 1=event, 2=competing risk) and a variable
indicating whether they started on treatment or not.

```{r, echo=FALSE}
# set working directory to TMLE_for_breakfast
baseline_data <- readRDS("Example_rtmle/baseline_data.rds")
treatment_data <- readRDS("Example_rtmle/treatment_data.rds")
head(baseline_data)
```

The other dataset has treatment periods for those allocated to
treatment. In the example treatment is numerical, but it may also be
dates.

```{r, echo=FALSE}
head(treatment_data)
```
# Prepare for LTMLE
LTMLE operates with a sequence of time periods all of equal length. Even though 
individuals may experience the event or be censored early, all individuals need
information for variables during the whole time of follow-up. Therefore, the first
step is to fix the end time of the analysis. 

For this example, we chose five periods corresponding to times of one to five

During the data preparation below, records are split into multiple records and 
to avoid confusion with the original entry into analysis, a new variables (inn/out)
are defined and used in further data management.

```{r, echo=TRUE}
base <- baseline_data[,.(ID,Trial_Start)]
base[,':='(inn=Trial_Start,out=Trial_Start+4*365)]
```
## Splitting
The following steps have the purpose of defining levels of variables in each of 
the defined time periods. To start this process all records are split according 
to timing of change in variable status. The order of splitting is not important.

The first step is splitting by all variables that change only once. 
This is performed with the heaven::splitTwo function that needs the original 
base data and a "splitting guide" which is the dataset with dates where variables
that change only once are held.  For the current example the only variables are censoring, outcome and competing risk nodes.  These variables are therefore defined in distinc variables. The list of variables can for realistic examples also include other time dependent variables.  Note that the content of each variable that is used for splitting needs to have value (numeric or date) at the time of change, otherwise NA.

```{r echo=TRUE}
splitguideTwo <- baseline_data[,.(ID,Trial_Start,Event_Type,Time)]
# use the multilevel event variable to create separate variable/time for censoring, outcome and competing risk
splitguideTwo[Event_Type==0,censor:=Time]
splitguideTwo[Event_Type==1,outcome:=Time]
splitguideTwo[Event_Type==2,compete:=Time]
longSplit <- splitTwo(indat=base,
                      splitdat=splitguideTwo,
                      invars=c('ID','inn','out'),
                      splitvars = c('ID','censor','outcome','compete'))
```

Next the data is split by the time dependent variables with potentially multiple
changes during the study. We use the function heaven::splitFromTo. A single call to the function can split on all time dependent variables representing intervals. Apart from variables indicating start and end of periods, two more variables are needed. One variable indicates a name for the condition and the other a value to assume when present. The function does not allow overlap whithin person/condition. This needs to be arranged prior to use of the function.

```{r echo=TRUE}
splitguideTime <- copy(treatment_data)
splitguideTime[,':='(treatment="treatment",value=1)]
longSplit <- splitFromTo(indat=longSplit,
                         splitdat=splitguideTime,
                         invars=c('ID','inn','out'),
                         splitvars = c('ID','Treatment_Start','Treatment_End','value','treatment'))

```

Finally, the data is split by the selected time periods, in the current example
just five periods. The new value 'period' contains the period number. This uses
heaven::splitSeq.

```{r echo=TRUE}
longSplit <- splitSeq(indat=longSplit,
                      invars=c('ID','inn','out'),
                      varname = 'Trial_Start', # intervals since inclusion in study
                      splitvector= seq(-1,5*365-1,365), # just five periods for the example
                      format = "vector",
                      value="period")
```
## Summarize in periods

With the splitting complete, all information for each selected equally
sized time period (in the example five) is separated.

The next step is then to summarize information by 'period'.

The outcomes (event, censoring, competing risk) should be the maximal
outcome for each period since an outcome event is coded "1" as opposed
to "0" when not occurring.

For other variables, it is dependent on a discussion whether to choose
the entry values for each period or choose any indication of exposure
during the period. For this example the first value is chosen for each
period for covariates.

For the calculations, the outcomes needs to come after
covariates. Depending on how the splitting is conducted then either
outcomes needs to move one period later or covariates one period
earlier.  The current version moves covariates.

```{r echo=TRUE}
setkeyv(longSplit,c("ID","period","inn"))
# Max value of outcomes in each period
longSplit[,':='(outcome=max(outcome),censor=max(censor),compete=max(compete)),
          by=c("ID","period")] 
setkeyv(longSplit,c("ID","inn"))
# Choose first record for each ID/period
aggrSplit <- longSplit[,.SD[1],by=c("ID","period")]
aggrSplit[,treatment:=as.numeric(treatment)]

# shift outcomes one period forward
aggrSplit_cov <- aggrSplit[,.(ID,period,treatment)] # all time dependent variables, here just "treatment"
aggrSplit_cov[,period:=period-1]
aggrSplit_out <- aggrSplit[,.(ID,period,outcome,censor,compete)] # outcomes

```
## Transpose to wide format
```{r echo=TRUE}
outcome_dt <- dcast(aggrSplit_out,ID~period,value.var = c("compete","censor","outcome"))
treatment_dt <- dcast(aggrSplit_cov,ID~period,value.var = c("treatment"))
# Correct column naming fails with just one variable:
setnames(treatment_dt,as.character(0:4),paste0("treatment_",0:4))
```
## LTMLE
```{r echo=TRUE}
x <- rtmle_init(intervals=5,name_id='ID',name_time='period',name_outcome='outcome',
    name_competing='compete',name_censoring='censor',
    censored_levels=c('1','0'),censored_label="1")
x$data <- list(baseline_data = baseline_data[,.(ID,Age)],
               outcome_data = outcome_dt,
               timevar_data = treatment_dt)
prepare_data(x) <- list()
protocol(x) <- list(name = "treat",treatment_variables = "treatment",intervention = 1)
protocol(x) <- list(name = "placebo",treatment_variables = "treatment",intervention = 0)
target(x) <- list(name = "Risk",strategy = "additive",estimator = "tmle",protocols = c("treat","placebo"))
x <- run_rtmle(x,time_horizon = 2)
summary(x)

```



