---
title: "LTMLE data preparation"
date: "19-05-2023"
---

This provides a simulated example to prepare what could be real data for a TMLE analysios.

All analyses are based on R and primarily using data.table for data management

The assumption is that the real raw data are found in several tables representing
diagnoses, personal data, prescriptions and outcomes.  So, for a starter we need to
create som simulated data for exercises representing such tables.

```{r}
library(heaven) # devtools::install_github("tagteam/heaven")
library(data.table)
library(ltmle)
set.seed(35)
```
# Step 1 - Create data for exercise

## Create a population
For some relation to reality we will create a population that has an age exceeding 50 years
in 2010.  
```{r}
n <- 100
population <- data.table(ID=1:n, birthDate=as.Date("2010-01-01")-(50*365+sample(1:6000,n)))
```
We assume that 20 percent die between 2010 and the following 4 years - our competing risk
```{r}
population[,dead:=sample(0:1,n,replace=TRUE,prob=c(0.8,0.2))]
population[dead==1,death:=as.Date("2010-01-01")+sample(1:1460,1,replace=FALSE)]
```
At baseline (2010-01-01) a single baseline variable is present - this is continuous and 
represents BMI
```{r}
population[,BMI:=sample(25:35,n,replace = TRUE)]
```

## Diagnoses representing outcome
We assume that a separate register provides diagnoses of patients in the study.
In reality the diagnoses would be ICD codes, but for ease we translate to diseases.
To match reality there are several (1,2) codes for some diseases. We introduce Myocardial
Infarction, Heart Failure and Renal Disease. The created data also contain various not of value for us.
```{r}
diagnoses <- simAdmissionData(n,m=5,c("MI","HF1","HF2","RD1","RD2"),startDate = "2000-01-01")
setnames(diagnoses,c("ID","someNumber","admission","discharge","otherDate","diagnosis","someNumber2"))
```

## Treatment with various drugs
We provide the possibility of GLP1 therapy to patients
We also provide the possibility of drugs with ATC codes C02A and C07FB  - which are helpful for 
identifying hypertension at least in Danish registers - providing two are used.
In this example patients can only develop hypertension after trial inclusion providing a time dependent 
comorbidity
```{r}
drugs <- simPrescriptionData(n,max.prescriptions = 50,packages=
          list("GLP1"=list(c(1,90)),"C02A"=list(c(1,90)),"C07FB"=list(c(1,90))),max.packages=2,
         startDate="2010-01-01")

setnames(drugs,c("ID","drug","prescriptionDate","strength","packageSize","numberPackages"))
```

# Step 2 - Arranging data

## Find comorbidity presence prior to entry
In this example all patients enter the study on "2010-01-01"
```{r}
morblist <- list(
  MI="MI",
  HF=c("HF1","HF2"),
  RD=c("RD1","RD2")
)
diagnoses2 <- findCondition(data=diagnoses,
                            vars="diagnosis",
                            keep=c("ID","admission"),
                            conditions=morblist,
                            match="start",
                            condition.name="morbidity")
setkeyv(diagnoses2,c("ID","morbidity","admission"))
priorDiagnoses <- diagnoses2[admission<as.Date("2010-01-01")]
priorDiagnoses <- priorDiagnoses[,.SD[1],by=c("ID","morbidity")]
priorDiagnoses[,presence:=fifelse(is.na(admission),0,1)]
priorDiagnoses <- dcast(priorDiagnoses,ID~morbidity,value.var="presence")
priorDiagnoses[is.na(priorDiagnoses),] <- 0  
# We assume that RD in the past is a different covariate than RD over time and give it a new name
setnames(priorDiagnoses,"RD","renal")
```
## Find date of comorbidity after entry
RD is considered a time dependent comorbidity
HF and AMI is a combined cardiovascular endpoint to be examined
```{r}
timeDependent <- diagnoses2[admission>as.Date("2010-01-01")]
timeDependent[morbidity %in% c("MI","HF"),morbidity:="outcome"]
setkeyv(timeDependent,c("ID","morbidity","admission"))
timeDependent <- timeDependent[,.SD[1],by=c("ID","morbidity")]
timeDependent <- dcast(timeDependent,ID~morbidity,value.var="admission")
```
## Find date of hypertension defined as the first quater where two
antihypertensives have been claimed for two consecutive quaters
```{r}
hypertension <- hypertensionMedication(data=drugs,
                                       vars=c("ID","drug","prescriptionDate")
                                       )
hypertension[,numDrugs:=NULL]
```
# Create the tables necessary for final data preparation
## ID and start date
```{r}
base <- data.table(ID=1:n,startDate=as.Date("2010-01-01"))
```
## Table of fixed variables at baseline
```{r}
fixed <- merge(population[,.(ID,birthDate)],priorDiagnoses,all.x=TRUE,by="ID")
fixed[,age:=as.numeric((as.Date("2010-01-01")-birthDate)/365.25)]
fixed[,birthDate:=NULL]
fixed[is.na(fixed),] <- 0  
```
## Table of variables that change only once - including outcome
At this stage we also add a censoring date among patients without outcome or competing risk

```{r}
timeDependent2 <- Reduce(function(x,y){merge(x,y,all=TRUE,by="ID")},
                        list(timeDependent,hypertension,population[,.(ID,death)]))
timeDependent2[is.na(death) & is.na(outcome),censor:=as.Date("2010-01-01")+sample(1:2000,1)]
```
For these simulated data we need an extra step because patients can only achive on of the
possibilities of censoring, outcome and death
```{r}
timeDependent2[!is.na(outcome),':='(censor=NA,death=NA)]
timeDependent2[!is.na(censor),death:=NA]
```



## Table of variables that can change multiple times - GLP1 treatment

In this case the simple decision is made that any prescription of GLP1 ensures
continuous treatment for 180 days

For use in following function overlap of treatment periods is not allowed within 
ID and drug.  Further the variable "value" is set to one representing that we in
this example regard treatment with GLP1 is qualitative 0/1
```{r}
timeVarCov <- drugs[drug=="GLP1",.(ID,drug,prescriptionDate)]
timeVarCov[,endDate:=prescriptionDate+180]
setkeyv(timeVarCov,c("ID","prescriptionDate"))
timeVarCov[,prescriptionDate:=fifelse(!is.na(shift(endDate)) & prescriptionDate<shift(endDate),shift(endDate),prescriptionDate)]
timeVarCov <- timeVarCov[endDate>prescriptionDate]
timeVarCov[,value:=1]
```

# Prepare for LTMLE
LTMLE operates with a sequence of time periods all of equal length. Even though 
individuals may experience the event or be censored early, all individuals need
information for variables during the whole time of follow-up. Therefore, the first
step is to fix the end time of the analysis. 

For this example, we chose four time intervals of 365 days.

During the data preparation below, records are split into multiple records and 
to avoid confusion with the original entry into analysis, a new variable is 
defined informing the start date.


```{r}
base[,':='(inn=startDate,out=startDate+4*365)]
```
## Splitting
The following steps have the purpose of defining levels of variables in each of 
the defined time periods. To start this proces all records are split according 
to timing of change in variable status. The order of splitting is not important.

The first step is splitting by all variables that change only once. 
This is performed with the heaven::splitTwo function that needs the original 
base data and a "splitting guide" which is the dataset with dates where variables
that change only once are held.

```{r echo=TRUE}
longSplit <- splitTwo(indat=base,
                      splitdat=timeDependent2,
                      invars=c('ID','inn','out'),
                      splitvars = c('ID','RD','outcome','hypertension','death','censor'))
```

Next the data is split by the time dependent variables with potentially multiple
changes during the study. We use the function heaven::splitFromTo.

```{r echo=TRUE}
longSplit <- splitFromTo(indat=longSplit,
                         splitdat=timeVarCov,
                         invars=c('ID','inn','out'),
                         splitvars = c('ID','prescriptionDate','endDate','value','drug'))
```

Finally, the data is split by the selected time periods, in the current case
four periods of 365 days. The new value 'period' contains the period number. This uses
heaven::splitSeq.

```{r echo=TRUE}
longSplit <- splitSeq(indat=longSplit,
                      invars=c('ID','inn','out'),
                      varname = 'startDate', # intervals since inclusion in study
                      splitvector= c(-1,365,730,1095,1460), # four periods of 180 days
                      format = "vector",
                      value="period")
```
## Summarize in periods
With the splitting complete, all information for each selected equally sized time period 
(in the example four) is separated.

The next step is then to summarize information by 'period'. 

The outcomes (event, censoring, competing risk) should be the maximal outcome for each period since an outcome event is coded "1" as opposed to "0" when not occurring.  

For other variables, it is dependent on a discussion whether to choose the entry values
for each period or choose any indication of exposure during the period. For this
example the entry value is chosen for each period for covariates. 

Importantly, the data structure for running LTMLE requires that when an event has
happened during a period, the value of that event needs to be carried forward to the
rest of periods until end of observation time.  

For the calculations, the covariates (as opposed to outcomes) need to be moved 
one period back in time. This ensures that covariate values are always prior to outcome. 
Conceptually covariate values can be viewed as the last value during the preceding time period.

```{r echo=TRUE}
setkeyv(longSplit,c("ID","period","inn"))
# Max value of outcomes in each period
longSplit[,':='(outcome=max(outcome),censor=max(censor),death=max(death)),by=c("ID","period")] 
setkeyv(longSplit,c("ID","inn"))
# Choose first record for each ID/period
aggrSplit <- longSplit[,.SD[1],by=c("ID","period")]
aggrSplit[,GLP1:=as.numeric(GLP1)]
# Make relation between censor, outcome and death possible - not relevant for REAL data

# shift covariates one period back 
aggrSplit_cov <- aggrSplit[,.(ID,period,GLP1,RD,hypertension)]
aggrSplit_cov[,period:=period-1]
aggrSplit_out <- aggrSplit[,.(ID,period,death,censor,outcome)]
```

## Transpose to wide format
```{r echo=TRUE}
wideSplit_cov <- dcast(aggrSplit_cov,ID~period,value.var = c("GLP1","RD","hypertension"))
wideSplit_out <- dcast(aggrSplit_out,ID~period,value.var = c("death","censor","outcome"))
wideSplit <- cbind(wideSplit_cov,wideSplit_out[,ID:=NULL])
# Finally add the baseline covariates
wideSplit <- merge(fixed,wideSplit,by="ID")
```
# Deterministic Q-function

The deterministic Q function can be employed for a variety of purposes. Centrally 
for our example, it interprets the "death" covariate as a competing risk.

```{r, echo=T}
det.Q.function<-function(data, current.node, nodes, called.from.estimate.g){
    compete.index <- grep("death",names(data))
    ## if(length(compete.index)==0) det <- list(rep(TRUE,NROW(data)),Q.value=0)
    ## stop("No compete/terminal event node found")
    hist.compete.index <- compete.index[compete.index < current.node]
    if(length(hist.compete.index)==0)
      return(NULL)
    else{
      is.deterministic <- Reduce("+",lapply(data[,hist.compete.index,drop=FALSE],
                                            function(dd){x=dd;x[is.na(dd)] <- 0;x}))>=1
      is.deterministic[is.na(is.deterministic)] <- FALSE
      list(is.deterministic=is.deterministic, Q.value=0)
    }
}
```

#LTMLE
What is left is to run the LTMLE analysis, and more information on this function exists: 
https://cran.r-project.org/web/packages/ltmle/ltmle.pdf

```{r echo=TRUE}
Anodes <- sort(grep("^GLP",names(wideSplit))) #treatment nodes
Lnodes <- sort(c(grep("HF",names(wideSplit)),grep("MI",names(wideSplit)),
                 grep("renal",names(wideSplit)),grep("^hypertension_",names(wideSplit)),
                 grep("^death",names(wideSplit)),grep("^RD_",names(wideSplit))))
Ynodes <- sort(grep("^outcome",names(wideSplit))) #outcome
Cnodes <- sort(grep("^censor",names(wideSplit))) #censor
    
fit <- ltmle(wideSplit,Anodes=Anodes, 
          Lnodes=Lnodes, 
          Ynodes=Ynodes,
          Cnodes=Cnodes,
          abar=c(1,1,1,1),
          deterministic.Q.function = det.Q.function, #defined above
          survivalOutcome = TRUE)

print(summary(fit))
```

